\large \textbf{Leveraging Historical Ties between Cognitive Science and Computer Science to Guide Programming Education}

### 2016-03-21 paper.tex
expand CLT section, connect to OOP

\titleformat*{\section}{\normalsize\bfseries} % Makes section titles 12 pt font
\titleformat{\section}{\normalsize\bfseries}{}{}{}[]
\titleformat{\subsection}{\bfseries}{}{}{}[]
\titleformat{\subsubsection}{\bfseries}{}{}{}[]

@@

Ultimately, if our goal is to teach problem solving in general, we
would be doing a grave disservice to our students to leave with the
impression that there is only one paradigm, whether it be OOP,
procedural, functional, or something not yet invented. As previously
mentioned, OOP and procedural paradigms are comprised of fundamentally
different concepts for problem solving, and so again is the functional
paradigm. Unsurprisingly, there is no paradigm that is the ``best''
fit for every problem. Leaving students with the belief that only one
paradigm exists makes it very difficult for them to learn how to
intentionally select and apply the ``best'' paradigm to solve a
particular problem. \citeauthor{kolling_problem_1999} argues that
performing a paradigm shift from procedural style to OOP is difficult,
and so concludes that if OOP is a desired outcome then it should be
taught \emph{first} since the reverse shift, from OOP to procedural,
is conceptually easier. What this

\subsection{Language choice}
Choice of language does matter: If the learning objectives of an
introductory programming course are to learn problem solving concepts
then a language with a simpler syntax is preferable to one with more
complex syntax features\autocite{koulouri_teaching_2014}. The
explanation for this can be found in cognitive load theory (CLT),
which posits that learners have a finite capacity for their short-term
memory, and for learning to occur, all aspects of the task must fit
into this finite space. While some contributors to cognitive load are
necessary for the task at hand, others are
not\autocite{paas_cognitive_2003}. If, for example, the goal is to
teach problem-solving techniques, there will be cognitive load
associated with thinking about different strategies one could
use. Learning the syntax rules of a programming language adds to the
cognitive load, but it does not aid in the process of learning
problem-solving techniques. Thus, it is in the educator's best
interest to minimize cognitive load associated with features that do
not directly help with the learning objectives of the course and
select a teaching language with easy-to-learn syntax.

@@

@@

\subsection{Cognitive Load Theory}

@@

If teaching OOP concepts is a goal of the course one must ask if
/implementation/ of OOP programs is also a goal.  If writing OOP code
is a goal and it is unreasonable to assume that students will have
already mastered syntax and enough understanding of the development
environment (interpreting compiler messages, debugging, etc) then CLT
would suggest that extensive /scaffolding/ be employed to reduce
extraneous cognitive load. Being careful to sequence simple-to-complex
concepts will help reduce /intrinsic/ cognitive load: i.e. teach bare
objects and make sure students have began to automate tasks associated
with implementing bare objects before introducing inheritance.

@@

The authors of the Lohani et al report describe efforts to use the
Alice programing environment to teach introductory programming
concepts\autocite{lohani_reformulating_2011}. In theory a graphical
programming environment should aid in the , learning of higher level
concepts since it eliminates the extraneous cognitive load of worrying
about syntax and grammar. In fact, there has been success in using
iconic languages to teach first year programming
\autocite{calloni_iconic_1997}, and research indicates that teaching
graphical programming first does not inhibit a student from later
learning textual programming languages\autocite{hundhausen_can_2009}.

Eric Raymond describes three different kinds of complexity in software
design: essential complexity, optional complexity, and accidental
complexity

>>>>

\titleformat*{\section}{\normalsize\bfseries} % Makes section titles 12 pt font
\titleformat{\section}{\normalsize\bfseries}{}{}{}[]
\titleformat{\subsection}[runin]{\bfseries}{}{}{}[.]
\titleformat{\subsubsection}[runin]{\itshape}{}{}{}[.]

@@

Ultimately, if our goal is to teach problem solving in general, we
would be doing a grave disservice to our students to leave with the
impression that there is only one paradigm, whether it be OOP,
procedural, functional, or something not yet invented. As previously
mentioned, OOP and procedural paradigms are comprised of fundamentally
different concepts for problem solving, and so again is the functional
paradigm. Unsurprisingly, there is no paradigm that is the ``best''
fit for every problem. Leaving students with the belief that only one
paradigm exists makes it very difficult for them to learn how to
intentionally select and apply the ``best'' paradigm to solve a
particular problem. \citeauthor{kolling_problem_1999} argues that
performing a paradigm shift from procedural style to OOP is difficult,
and so concludes that if OOP is a desired outcome then it should be
taught \emph{first} since the reverse shift, from OOP to procedural,
is conceptually easier. What this recommendation neglects to account
for however is that both paradigms depend upon authoring programs,
whether they be textual or iconic, and this authoring process takes a
certain amount of cognitive capacity.

\subsection{Cognitive load theory}
Cognitive load theory (CLT) is based on the premise that the capacity
of Working Memory, where all conscious cognitive processing takes
place, has a finite and relatively small capacity: we are able to
manage 2-3 interacting elements at a time in our working
memory. Examples of ``interacting elements'' in the current context
would be something like recalling and using the syntax and grammar of
a particular programming language, reasoning about the behavior of a
single object in an OOP setting, or reasoning about a solution to a
particular high-level problem. CLT posits that learners have a finite
capacity for their short-term memory, and for learning to occur, all
aspects of the task must fit into this finite space. While some
contributors to cognitive load are necessary for the task at hand,
others are not\autocite{paas_cognitive_2003}.

In a superficial way, CLT theory would suggest that OOP is a helpful
paradigm for managing complexity: objects encapsulate complexity by
hiding the details of implementation behind a simple interface. For
example, and object that represents a list of items may have
operations to add and remove items from the list. Conceptually we can
use this object with our common understanding of what it means to
``add'' and ``item'' to a ``list''. Critically, we do \emph{not} need
to be aware of how the adding and removal of the item is actually
implemented in code, or even whether how the storage is structured in
memory (e.g. an array or a linked-list). We refer to this use of CLT
to justify the reduction of complexity of OOP as superficial because
it focus on the conceptual paradigm itself without regard to
implementing it in some programming environment. In practice, we
usually ask our students to implement OOP concepts in a program. That
task requires not only understanding OOP as a paradigm, but also
working with the programming environment, authoring a program, and
most likely debugging it as well.

\subsection{Language choice}
Choice of language does matter: If the learning objectives of an
introductory programming course are to learn problem solving concepts
then a language with a simpler syntax is preferable to one with more
complex syntax features\autocite{koulouri_teaching_2014}. The
explanation for this can be found in CLT. If, for example, the goal is to
teach problem-solving techniques, there will be cognitive load
associated with thinking about different strategies one could
use. Learning the syntax rules of a programming language adds to the
cognitive load, but it does not aid in the process of learning
problem-solving techniques. Thus, it is in the educator's best
interest to minimize cognitive load associated with features that do
not directly help with the learning objectives of the course and
select a teaching language with easy-to-learn syntax.

@@

\subsubsection{Iconic versus textual languages}

In theory a iconic programming environment such as
Scratch\autocite{resnick_scratch_2009} or
Alice\autocite{cooper_using_2003} should aid in the learning of higher
level concepts since it eliminates the extraneous cognitive load of
dealing with syntax and grammar of a language. In fact, there has been
success in using iconic languages to teach first year programming
\autocite{calloni_iconic_1997}, and research indicates that teaching
graphical programming first does not inhibit a student from later
learning textual programming languages\autocite{hundhausen_can_2009}.

\citeauthor{snook_incorporation_2005} describe efforts to use the Alice
programing environment to teach introductory programming
concepts\autocite{snook_incorporation_2005}. Interestingly a later
report assessing the efficacy of the curriculum which included this
use of Alice indicated that while pre-post testing indicated learning
gains with Alice, focus group data indicated a dissatisfaction with
Alice as an introductory language which was later replaced with
LabVIEW\autocite{lohani_reformulating_2011}.

@@

@@

If teaching OOP concepts is a goal of the course one must ask if
\emph{implementation} of OOP programs is also a goal.  If writing OOP
code is a goal and it is unreasonable to assume that students will
have already mastered syntax and enough understanding of the
development environment (interpreting compiler messages, debugging,
etc) then CLT would suggest that extensive \emph{scaffolding} be
employed to reduce extraneous cognitive load. Being careful to
sequence simple-to-complex concepts will help reduce /intrinsic/
cognitive load: i.e. teach bare objects and make sure students have
began to automate tasks associated with implementing bare objects
before introducing inheritance.

@@

Eric Raymond describes three different kinds of complexity in software
design: essential complexity, optional complexity, and accidental
complexity. This has interesting parallels to the different types of
cognitive load in CLT. In particular the notion of \emph{essential
  complexity}, like \emph{intrinsic} cognitive load is associated with
the the solution itself: a solution to a complex problem will be
essentially complex due to the complexity associated with the problem
itself.
